---
title: "Statistical and Data Analysis"
subtitle: "for the (social) data scientist"
author: "Austin Hart, Ph.D."
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, nhsr, nhsr-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
---


```{r setup, include = FALSE}
## Libraries
  library(knitr)
  library(rmdformats)

## Global options
  knitr::opts_chunk$set(
    echo = TRUE, 
    prompt = FALSE,
    tidy = FALSE, 
    comment = NA,
    message = FALSE, 
    warning = FALSE,
    dev = 'CairoSVG',
    fig.align = 'center'
)
```

# Following along

- Fork and clone [AnalysisOverview repo](https://github.com/austin-hart-pols/AnalysisOverview)

- `backloash.dta`: 2016 ANES Survey + 2020 re-interview; white (non-Hispanic) subsample

```{r}
  library(tidyverse); library(haven); library(janitor); library(knitr)

## Data
  df = read_dta('backlash.dta') |>
    mutate(
      pid2016 = case_when(
        PartyID2016 %in% 1:3 ~ '1. DEM',
        PartyID2016 == 4 ~ '2. IND',
        PartyID2016 %in% 5:7 ~ '3. GOP'
      )
    ) |>
    mutate(across(where(is.labelled), ~ as_factor(.)))
```


---
class: inverse, middle

# Counts and frequencies
### For categorical variables

- One-way tables  

- Two-way (contingency) tables

- Test of independence: $\chi^2$  


---
# One-way tables

> Describe distribution of scores on a single, categorical variable.

.pull-left[
### `tidyverse` approach

```{r}
  df |>
    count(pid2016) |>
    na.omit() |>
    mutate(percent = 100 * n / sum(n)) |>
    kable(digits = 1L)
```
]

.pull-right[
### `janitor` alternative

```{r}
  df |>
    tabyl(pid2016, show_na = F) |>
    adorn_totals() |>
    adorn_pct_formatting() |>
    kable(digits = 1L)
```
]

---
# Significant variation?

.pull-left[

### Obs vs Exp
```{r, echo = F}
  df |>
    count(pid2016) |>
    na.omit() |>
    mutate(per = 100 * n / sum(n)) |>
    mutate(
      nExp = 0.5 * sum(n),
      perExp = 100 * nExp / sum(nExp)
    ) |>
    kable(digits = 1L)
```

Is the obs freq significantly different from expected? Find the probability we observe this by chance alone.

Yes, there is significant variation (unequal proportions) across categories $(\chi^2_2 = 640, ~ p < 0.001)$

]

.pull-right[
### Goodness of Fit, $\chi^2$
```{r}
  df |>
    tabyl(pid2016, show_na = F) |>
    pull(n) |>
    stats::chisq.test()
```
]

---
# Contingency (two-way) tables

Describe joint, marginal distribution for two categorical variables.

### Code

```{r, eval = FALSE}
## Raw cross-tab
  t1 = 
    df |>
    tabyl(
      pid2016, Female, # Y, X
      show_na = F,
      show_missing_levels = F
    ) 

## Pretty table
  t1 |>
    adorn_totals(where = c("row", "col")) |>
    adorn_percentages('col') |>
    adorn_pct_formatting(digits = 1) |>
    adorn_title()
```

---
# Contingency (two-way) tables

Describe joint, marginal distribution for two categorical variables.

.pull-left[
### Table Output

```{r, echo = FALSE}
## Raw cross-tab
  t1 = 
    df |>
    tabyl(
      pid2016, Female, # Y, X
      show_na = F,
      show_missing_levels = F
    ) 

## Pretty table
  t1 |>
    adorn_totals(where = c("row", "col")) |>
    adorn_percentages('col') |>
    adorn_pct_formatting(digits = 1) |>
    adorn_title() |>
    kable()
```
]

.pull-right[
### Interpretation
Read down the columns (i.e., within categories of X). 

Among females, 42% identify as Democrats, 45% as Republicans, and about 12% Independent. Among non-females, over half (53%) identify as Republicans, about one third (36%) Democrats, and 12% Indpendent.
]

---
# Contingency (two-way) tables

Describe joint, marginal distribution for two categorical variables.

.pull-left[
### $\chi^2$ Test of independence

Does the distribution of outcomes depend on (i.e., change with) the value of the exposure variable?

```{r}
  chisq.test(t1)
```

]

.pull-right[
Yes. There is evidence of a systematic relationship between party identification and sex.

It is unlikely we observe a pattern like this by chance alone $(\chi^2_2 = 18.9,~p<0.001)$.

*Remember* that the $p-value$ indicates the probability of your sample data given the assumption that party ID and sex are independent (unrelated).
]

---
class: inverse, middle

# Summary statistics
### For numeric variables

- One variable 

- Group comparisons

- $t$-tests

- skew and the $log$-transformation


---
# Describing Distributions
### What to show and tell

- Give context (min to max)

- Cite location (mean)

- Describe variation (standard deviation)

- Describe shape (skew)

- Present a graph


---
# Single continuous variable
### Gather info

.pull-left[
```{r}
## useful stats
  summary(df$ImmIndex)
  sd(df$ImmIndex, na.rm = T)
```
]

.pull-right[
```{r, fig.dim=c(4,3)}
## plots
  hist(df$ImmIndex, main = NULL)
```
]

---
# Immigration Support
### Among 2016 ANES respondents (white, non-Hispanic)

.pull-left[
- Creating the index score
  - Combines responses from 7 questions

  - High scores = pro-immigration  
  
- Scores in 2016 (N = 3,038)

  - *Range of values:* `r min(df$ImmIndex, na.rm = T) |> round(2)` to `r max(df$ImmIndex, na.rm = T) |> round(2)`

  - *Mean*: `r mean(df$ImmIndex, na.rm = T) |> round(2)`

  - *Std Dev*: `r sd(df$ImmIndex, na.rm = T) |> round(2)`

]

.pull-right[

```{r, echo=FALSE, fig.dim=c(4,4)}
df |>
  ggplot(aes(x = ImmIndex)) +
  stat_density(trim = F, fill = 'cornflowerblue') +
  geom_histogram(aes(y = ..density..), color = 'black', fill = NA, bins = 15) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(y = 'Frequency', x = 'Immigration index score') +
  theme_classic(base_size = 14) +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

```

]

---
# Comparing groups
### Subgroup (mean) comparison

.pull-left[
### Subgroup code

```{r, eval = FALSE}
## Group means
  df |>
    group_by(Female) |>
    summarise(
      Avg = mean(ImmIndex, na.rm = T),
      SD = sd(ImmIndex, na.rm = T)
    ) |>
    kable(digits = 2L)

## Compare distributions
  df |>
    ggplot(aes(x = ImmIndex, fill = Female)) +
    stat_density(
      position = 'dodge', 
      alpha = 0.3, color = 'black'
    )
```
]

.pull-right[
### Outputs

```{r, echo = F}
  df |>
    group_by(Female) |>
    summarise(
      Avg = mean(ImmIndex, na.rm = T),
      SD = sd(ImmIndex, na.rm = T)
    ) |>
    kable(digits = 2L)
```

```{r, echo = F, fig.dim=c(4,2.5)}
  df |>
    ggplot(aes(x = ImmIndex, fill = Female)) +
    stat_density(
      position = 'dodge', 
      alpha = 0.3, color = 'black'
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    theme_classic(base_size = 14)
```
]

---
# Mean-focused alternative

.pull-left[
```{r, eval = FALSE}
## Group means
  s1 = df |>
    group_by(pid2016) |>
    summarise(
      Avg = mean(ImmIndex, na.rm =T)
    ) |>
    na.omit()
  
## Plotting
  s1 |>
    ggplot(aes(x = pid2016, y = Avg, 
               color = pid2016)) +
    geom_hline(yintercept = 0) +
    geom_point(size = 3) +
    geom_segment(aes(
      x = pid2016, xend = pid2016, 
      y = 0, yend = Avg
    ))
```
]

.pull-right[
```{r, echo = FALSE, fig.dim=c(4,4)}
## Group means
  df |>
    group_by(pid2016) |>
    summarise(
      Avg = mean(ImmIndex, na.rm =T)
    ) |>
    na.omit() |>
    ggplot(aes(x = pid2016, y = Avg,
               color = fct_rev(pid2016),
               label = round(Avg, 2))) +
    geom_hline(yintercept = 0) +
    geom_point(size = 4) +
    geom_text(hjust = 0, nudge_x = 0.11) +
    geom_segment(
      aes(
        x = pid2016, xend = pid2016, 
        y = 0, yend = Avg
      ),
      size = 1
    ) +
    labs(
      x = NULL,
      y = 'Mean Immig Support'
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.text.y = element_blank(),
      legend.position = 'none'
    )
```
]


---
# Testing for differences

.pull-left[
### Diff of (2) means

```{r}
  t.test(ImmIndex ~ Female, df)
```
Support for immigration does not differ systematically for females vs non-females $(t = -0.11, ~ p = 0.91)$.

]

.pull-right[
### ANOVA for more means
```{r}
  summary(aov(ImmIndex ~ pid2016, df))
```


Attitudes about immigration differ significantly by party identification. It is unlikely we observe these differences by chance alone $(F=364, ~p < 0.001)$.

]

---
# What about skew?

.pull-left[
- The Problem   

  - Skewness is the pull of the mean from the median

  - Created by one-sided extreme values

  - Renders mean a poor measure of location/middle

- Solution  

  - Transform the data to mitigate skew
  
  - For positive skew: use $ln(X)$ or $log_{10}(X)$
]

.pull-right[
```{r, echo = FALSE, fig.dim=c(4,4.5)}
  GDP = gapminder::gapminder |>
    filter(year == 1997) |> 
    mutate(
      country = country, 
      gdp = gdpPercap,
      metric = 'Original',
      .keep = 'none'
    )
  G2 = gapminder::gapminder |>
    filter(year == 1997) |> 
    mutate(
      country = country, 
      gdp = log(gdpPercap),
      metric = 'Logged',
      .keep = 'none'
    )
  
  bind_rows(GDP, G2) |>
    ggplot(aes(x = gdp)) +
    facet_wrap(~  fct_rev(metric), nrow = 2, 
               scales = 'free_x', strip.position = 'right') +
    geom_histogram(bins = 15, color = 'white') +
    scale_y_continuous(expand = expansion(mult = c(0, 0.4))) +
    labs(
      y = NULL,
      x = 'Measure of GDP'
    ) +
    theme_classic(base_size = 14) +
    theme(
      axis.line.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )
    
```
]

---
class: inverse, middle

# Linear relationships
### For many numeric variables

- Correlation

- OLS regression

- Regression with panel data


---
# Linear Correlation

.pull-left[
### Correlation coefficient, $r$

- measure of linear dependence 

- indexes direction and magnitude

- useful for exploring relationships

]

.pull-right[
```{r}
  gapminder::gapminder |>
    select(where(is.numeric), -year) |>
    corrr::correlate() |>
    kable(digits = 2L)
```
]

---
# Regression: Basics
